---
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(forecast)
library(tidyverse)
```

```{r}
#singapore 25-54 labour force participation rate (%)
# data obtained from gapminder.com/data

#70.19999695	71.09999847	71.69999695	72.30000305	73.09999847	72.90000153	74.40000153	
# 75.40000153	76	76.90000153	77.40000153	78.19999695	78.90000153	78.69999695	79	78	
# 80	79.80000305	80.30000305	80.5 80.69999695	81.09999847	81.5	81.90000153	
# 82.30000305	82.69999695	83.19999695	83.5

#import with scan
singapore <- c(70.19999695,	71.09999847,	71.69999695,	72.30000305,	73.09999847,	72.90000153,	74.40000153,	75.40000153,	76,	76.90000153,	77.40000153,	78.19999695,	78.90000153,	78.69999695,	79,	78,	80,	79.80000305,	80.30000305,	80.5,	80.69999695,	81.09999847,	81.5,	81.90000153,	82.30000305,	82.69999695,	83.19999695,	83.5)

#conversion to time series
singapore <- ts(singapore, start = 1980)

plot(singapore, ylab = "Labour force participation rate 25-54 %")
```
```{r}
singapore
```


 Exponential smoothing with holt
 
```{r}
holttrend <- holt(singapore, h = 5) 
summary(holttrend)
```

The trend

```{r}
plot(holttrend)
```

The trend can't continue beyond 100%. 
As such, we need to dampen the model.

```{r}
holt(singapore, damped = TRUE, h = 15) %>%
  plot()
```

```{r}
holt(singapore, damped = TRUE, h = 15) %>%
  summary()
```

```{r}
holt(singapore, damped = TRUE, phi=0.8, h = 15) %>%
  plot()
```

## The ARIMA model

Since the data is trending, an Auto correlation exists. Auto correlation means an earlier observation influences the later observations. This is what happens in trending data.

First parameter of ARIMA model is AR. A data set containing a clear trend will mostly consist of an AR model. Moving Average(MA) is mostly present in random or flat datasets. Since this dataset is not seasonal, we don't need the seasonal model.

We'll start with auto.arima function from forecast package. This will lead to automatic parameter selection. 

```{r}
singaporearima <- auto.arima(singapore)
summary(singaporearima)
```

Beecause there is a trend in the dataset, R recognized that and has AR part set to 1. 
The model also chose first order of differencing. R uses the differences between observations for modelling. 
Drift gives the bias added to the model.

```{r}
forecast(singaporearima, h=5) %>%
  plot()
```

Exact calculation of parameters

```{r}
auto.arima(singapore, stepwise = FALSE, approximation = FALSE)
```

```{r}
holttrend <- holt(singapore, h=10)
holtdamped <- holt(singapore, damped = TRUE, h=10)
arimaauto <- auto.arima(singapore) %>%
  forecast(h=10)
```

```{r}
autoplot(singapore) +
  autolayer(holttrend$mean, series="Holt Linear Trend") +
  autolayer(holtdamped$mean, series="Holt Damped Trend") +
  autolayer(arimaauto$mean, series="ARIMA") +
  xlab("year") + ylab("Labour Force Participation Rate Age 25-54") +
  guides(color=guide_legend(title = "Forecast Method") +
           theme(legend.position = c(0.8, 0.2))) +
  ggtitle("Singapore") + 
  theme(plot.title = element_text(family = "Times", hjust = 0.5,
                                  colour = "blue", face="bold",
                                  size = 15))
```

```{r}
autoplot(singapore) + geom_line(size=2)+
  autolayer(holttrend$fitted, series="Holt Linear Trend", size = 1.1) +
  autolayer(holtdamped$fitted, series="Holt Damped Trend", size = 1.1) +
  autolayer(arimaauto$fitted, series="ARIMA", size = 1.1) +
  xlab("year") + ylab("Labour Force Participation Rate Age 25-54") +
  guides(color=guide_legend(title = "Forecast Method") +
           theme(legend.position = c(0.8, 0.2))) +
  ggtitle("Singapore") + 
  theme(plot.title = element_text(family = "Times", hjust = 0.5,
                                  colour = "blue", face="bold",
                                  size = 15))
```


## US Annual Employment Count

```{r}
employment_count <- read.csv('~/Downloads/annual-employment-us-1890-to-197.csv', nrows = 81)
```

```{r}
head(employment_count)
```


```{r}
#conversion to time series
employment_count <- ts(employment_count$Annual.employment..U.S...1890.to.1970, 
                       start = 1890, end = 1970)

plot(employment_count)
```

```{r}
holttrend <- holt(employment_count, h = 10) 
summary(holttrend)
```

The trend

```{r}
plot(holttrend)
```

The trend can't continue beyond 100%. 
As such, we need to dampen the model.

```{r}
holt(employment_count, damped = TRUE, h = 15) %>%
  plot()
```

```{r}
holt(employment_count, damped = TRUE, h = 15) %>%
  summary()
```


## The ARIMA model

Since the data is trending, an Auto correlation exists. Auto correlation means an earlier observation influences the later observations. This is what happens in trending data.

First parameter of ARIMA model is AR. A data set containing a clear trend will mostly consist of an AR model. Moving Average(MA) is mostly present in random or flat datasets. Since this dataset is not seasonal, we don't need the seasonal model.

We'll start with auto.arima function from forecast package. This will lead to automatic parameter selection. 

```{r}
employmentarima <- auto.arima(employment_count)
summary(employmentarima)
```

Beecause there is a trend in the dataset, R recognized that and has AR part set to 1. 
The model also chose first order of differencing. R uses the differences between observations for modelling. 
Drift gives the bias added to the model.

```{r}
forecast(employmentarima, h=10) %>%
  plot()
```

Exact calculation of parameters

```{r}
auto.arima(employment_count, stepwise = FALSE, approximation = FALSE)
```

```{r}
holttrend <- holt(employment_count, h=60)
holtdamped <- holt(employment_count, damped = TRUE, h=60)
arimaauto <- auto.arima(employment_count) %>%
  forecast(h=60)
```

```{r}
autoplot(employment_count) +
  autolayer(holttrend$mean, series="Holt Linear Trend") +
  autolayer(holtdamped$mean, series="Holt Damped Trend") +
  autolayer(arimaauto$mean, series="ARIMA") +
  xlab("year") + ylab("Labour Force Participation Rate Age 25-54") +
  guides(color=guide_legend(title = "Forecast Method") +
           theme(legend.position = c(0.8, 0.2))) +
  ggtitle("Singapore") + 
  theme(plot.title = element_text(family = "Times", hjust = 0.5,
                                  colour = "blue", face="bold",
                                  size = 15))
```

```{r}
autoplot(employment_count) + geom_line(size=2)+
  autolayer(holttrend$fitted, series="Holt Linear Trend", size = 1.1) +
  autolayer(holtdamped$fitted, series="Holt Damped Trend", size = 1.1) +
  autolayer(arimaauto$fitted, series="ARIMA", size = 1.1) +
  xlab("year") + ylab("Labour Force Participation Rate Age 25-54") +
  guides(color=guide_legend(title = "Forecast Method") +
           theme(legend.position = c(0.8, 0.2))) +
  ggtitle("Singapore") + 
  theme(plot.title = element_text(family = "Times", hjust = 0.5,
                                  colour = "blue", face="bold",
                                  size = 15))
```